---
title: "PhenoGeneRanker"
author: "Cagatay Dursun, Jake Petrie,  Serdar Bozdag"
date: "7/15/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PhenoGeneRanker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PhenoGeneRankerPackage)
```
# Table of Contents
  * [Introduction](#Introduction)
  * [Using the Methods](#Using-the-Methods)  
    *[Input File Formatting](#Input-File-Formatting)  
    *[CreateWalkMatrix](#CreateWalkMatrix)  
    *[RandomWalkRestarts](#RandomWalkRestarts)  
  * [Example Workflow](#Example-Workflow)
  * [Summary](#Summary)
  * [Acknowledgements](#Acknowledgements)
  
## <a id = "Introduction"></a> Introduction

PhenoGeneRanker is a gene/disease prioritization tool that utilizes heterogeneous gene disease network. PhenoGeneRanker allows multi-layer gene and disease networks. It also calculates empirical p-values of gene ranking using random stratified sampling of genes based on their connectivity degree in the network.

# Load PhenoGeneRanker

Loading the package will make it usable and check to make sure that all of the nessecary packages are also installed. 

```{r}
library(PhenoGeneRankerPackage)
```

## <a id = "Using-the-Methods"></a> Using the Methods

* CreateWalkMatrix
* RandomWalkRestarts


## <a id = "Input-File-Formatting"></a> Input File Formatting
The CreateWalkMatrix method takes in an input file as a parameter. This file is a ".txt" file with two tab-separated columns which hold the data files you would like to work with. The header row must say “type” and “file_name” for each of the two columns. The type column contains what kind of data is stored in the file. These can be either gene, phenotype, or bipartite file types. The file_name column stores the name of your data file along with the “.txt” extension.   

This file should look something like this:
```
type    file_name  
gene    geneFile1.txt 
gene    geneFile2.txt
phenotype    phenotypeFile1.txt  
phenotype    phenotypeFile2.txt
bipartite    bipartiteFile.txt  
```

Inside each file_name files, there is a “from”, “to”, and “weight” column which are all tab-separated. For the gene and phenotype files, the “from” and “to” columns will contain genes and phenotypes, the row order doesn’t matter. The “weight” column will have the weight of the relationship of the genes and phenotypes, for unweighted network all weights must have a value of 1. For the bipartite file, the “from” column must have genes, the “to” column must have phenotypes, the meaning and usage of weight is similar to the gene and phenotype layers.  

The geneFile1.txt from the previous example could look like this:
```
from     to       weight  
gene1    gene2    1  
gene1    gene3    1  
gene2    gene4    1 
...      ...      ...
```
The phenotype files could look like this:
```
from     to       weight  
pheno1   pheno2   1  
pheno1   pheno3   1  
pheno3   pheno5   1  
```
The bipartite file would look like this:
```
from     to        weight  
gene1    pheno1    1  
gene2    pheno2    1  
```

## <a id = "CreateWalkMatrix"></a> CreateWalkMatrix

This method Generates a Walk matrix from Gene and Phenotype data. All it needs to take in as a parameter is an input file. It has to be a '.txt' file. This text file is made from two columns that are tab-separated. The header row needs the two columns to say type and file_name. Every row after that contains what kind of data each file would contain. These can be either gene, phenotype, or bipartite files. Under file_name, you will input the name of the file along with the .txt extension. Inside each file_name files, there is a from, to, and weight column which are all tab-separated. For gene and phenotype layer files, the “from” and “to” columns will have genes and phenotypes, the order doesn’t matter. The “weight” column will have the weight of the relationship of the genes and phenotypes, for unweighted network all weights must have a value of 1. For a bipartite file, the from column must have genes, the “to” column must have phenotypes, the meaning and usage of weight is similar to the gene and phenotype layers.  

The rest of the parameters are optional, leave them blank if you would like to use their default values. The parameter numCores is the number of cores used for parallel processing, it defaults to 1. It is recommended to use detectCores() for this parameter. The parameter delta is the probability of jumping between gene layers. It has a range of 0 to 1. It has a default value of 0.5. The parameter zeta is the probability of jumping between gene layers. It has a range of 0 to 1. It has a default value of 0.5. The parameter lambda is the Inter-Network Jump Probability. It has a range of 0 to 1. It has a default value of 0.5.   

The returned list is what will be used in the RandomWalkRestarts() method.

```{r eval=FALSE}
walkMatrix <-CreateWalkMatrix('file.txt')
```

If you would like to access the individual elements of this list you could do so by typing this.
```{r eval=FALSE}
walkMatrix[[“WM”]] # accesses the walk matrix itself
walkMatrix[[“genes”]] # accesses the sorted gene pool nodes
walkMatrix[[“phenotypes”]] # accesses the sorted phenotype pool nodes
walkMatrix[[“gene_connectivity”]] # accesses the degree of all the genes in the network
walkMatrix[[“phenotype_connectivity”]] # accesses the degree of all the phenotypes in the network
walkMatrix[[“LG”]] # the number of genes in the network
walkMatrix[[“LP”]] # the number of phenotypes in the network
walkMatrix[[“N”]] # the number of gene pool nodes
walkMatrix[[“M”]] # the number of phenotype pool nodes
```


## <a id = "RandomWalkRestarts"></a> RandomWalkRestarts

This method runs the random walk with restarts on the provided walkmatrix transition matrix of multiplex heterogeneous networks. When random walk converges to steady state then the random walk stops. If the parameter generatePvalue is FALSE, it returns a data frame including gene and phenotype names, ranks and steady state scores of the genes and phenotypes. If generatePvalue is TRUE then it generates p-values along with the ranks with respect to offset value of 100.

```{r eval=FALSE}
rwr<-RandomWalkRestarts(walkMatrix, c('gene1', 'gene2'), c(), TRUE)
```

If the generatePvalues parameter is TRUE, the resulting data frame looks like this.
```
gene    score     P100
g1      1.9e-3    0.020000000
g425    1.6e-3    0.030000000
g77     1.3e-3    0.120000000
...     ...       ...
```
If the generatePvalues parameter is FALSE, the resulting data frame looks like this.
```
rank    gene    score
1       gene47  8.877424e-04
2       gene202 8.219792e-04
3       gene33  5.779120e-04
...     ...     ...
```

Due to the nature of this method, it can return a different output depending on the "generatePvalue" parameter. If the parameter "generatePValue" is TRUE, then this method returns a rank sorted data frame with three columns for the gene name, score, p-value. If "generatePValue" is FALSE, then it creates a data frame of all of the genes ranked. It has a column of the rank, gene name, and random walk scores for each gene or phenotype.

## <a id = "Example-Workflow"></a> Example Workflow
It is recommended that you should run these two methods one after another.  
```{r eval=FALSE}
walkMatrix <-CreateWalkMatrix('file.txt')
rwr <-RandomWalkRestarts(walkMatrix, c('gene1', 'gene2'), c(), TRUE)
```

## <a id = "Summary"></a> Summary
This package was originally created for a gene prioritization project. It was performed with datasets from rice genes. You can read the paper about this project here. [PhenoGeneRankerPaper](https://www.biorxiv.org/content/10.1101/651000v1)

## <a id = "Acknowledgements"></a> Acknowledgements
package specific
